Draft Exploratory Analysis

### 2024-12-02 Data_Cleaning_20241031.Rmd
Add files via upload

Load Raw Data/Packages: *Data loaded from path on my local computer
```{r}
library(readr)
library(tidyverse)
library(readxl)
library(sf)
library(tmap)
library(raster)
library(rvest)
library(units)
library(dplyr)
library(ggplot2)

setwd("C:/Users/kdfro/Documents/MSU/ECNS 560/Term Project")
Bison_numbers <- read_csv("Data/Bison_numbers_by_the_county National Bison Association.csv")
hunting_data_tidy <- read_excel("Data/hunting_data_tidy.xlsx")
```

@@

#Load Habitats:
habitats = raster("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/USGS Bison Habitat Map/mBISOx_CONUS_HabMap_2001v1/mBISOx_CONUS_HabMap_2001v1.tif")

@@

Quick Spatial Plot:
```{r}
qtm(ranges)
qtm(habitats)
```

@@

graph2_data = merged_bison_data |>
  filter(state_code == "MONTANA" | state_code == "WYOMING" | state_code == "NORTH DAKOTA" | state_code == "SOUTH DAKOTA" | state_code == "NEBRASKA" | state_code == "KANSAS" | state_code == "OKLAHOMA" | state_code == "TEXAS" | state_code == "IDAHO" | state_code == "COLORADO" | state_code == "NEW MEXICO")

@@

@@

#Overall trends across the great plains... see downward trend overall
ggplot(graph2_data, aes(x = year, y = total_population)) + 
  xlab("Year") +
  ylab("Population") +
  labs(title = "Bison Population Over Time") +
  geom_point(aes(color = state_code), size = 3) +
  geom_smooth()

#Number of Hunting licenses Over Time
ggplot(graph2_data, aes(x = year, y = `Paid Hunting License Holders`)) + 
  xlab("Year") +
  ylab("Population") +
  labs(title = "Number of Licenses Over Time") +
  geom_point(aes(color = state_code), size = 3) +
  geom_smooth()

#Looking for relationship between number of hunting licenses and population
ggplot(merged_bison_data, aes(x = `Paid Hunting License Holders`, y = total_population)) +
  geom_point() + 
  geom_smooth() +
  xlab("Number of Lisense Holders") +
  ylab("Population") +
  labs(title = "Number of Licenses vs Population")
```

>>>>

Load Raw Data/Packages: *Data loaded from path on my local computer
```{r}
library(readr)
library(tidyverse)
library(readxl)
library(sf)
library(tmap)
library(raster)
library(rvest)
library(units)
library(dplyr)
library(ggplot2)
library(units)

Bison_numbers <- read_csv("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/Bison_numbers_by_the_county National Bison Association.csv")
hunting_data_tidy <- read_excel("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/hunting_data_tidy.xlsx")
```

@@

@@

#Load States:
states = st_read("C:/Users/kdfro/Documents/MSU/GitHub/course-materials/lecture-slides/12-Spatial/data/us_states_contiguous/states_contiguous.shp")
```
Merge States and Ranges:
```{r}
#Check CRS
st_crs(states)
st_crs(ranges)

#Transform/Project:
ranges2 = st_transform(ranges, crs = st_crs(states))
stopifnot(st_crs(ranges2) == (st_crs(states)))

tm_shape(states) + tm_polygons() +
  tm_shape(ranges2) + tm_polygons(col = "darkgreen")
```

Spatial Cleaning/Exploration:
```{r}
#Split ranges df into multiple polygons
ranges3 <- st_cast(ranges2, "POLYGON")

#Create variable for area
ranges3$area = st_area(ranges3)

#Change units from m^2 to km^2
set_units(st_area(ranges3), km^2)

#Visualize
tm_shape(ranges3) + tm_polygons(col = "area")

#Select states of interest:
states2 = states |>
  filter(STATE == "Colorado" | STATE == "Idaho" | STATE == "Kansas" | STATE == "Montana" | STATE == "Nebraska" | STATE == "New Mexico" | STATE == "North Dakota" | STATE == "North Dakota" | STATE == "Oklahoma" | STATE == "South Dakota" | STATE == "Texas" | STATE == "Wyoming")

#Find intersection of ranges and states:
ranges_intersect = st_filter(ranges3, states2)

#Visualize:
tm_shape(states2) + tm_polygons() +
  tm_shape(ranges_intersect) + tm_polygons(col = "darkgreen")
```

@@

graph2_data = merged_bison_data |>
  filter(state_code == "MONTANA" | state_code == "WYOMING" | state_code == "NORTH DAKOTA" | state_code == "SOUTH DAKOTA" | state_code == "NEBRASKA" | state_code == "KANSAS" | state_code == "OKLAHOMA" | state_code == "IDAHO" | state_code == "COLORADO" | state_code == "NEW MEXICO") |>
  mutate(population_bin = cut(total_population, breaks = c(0, 2000, 5000, 10000), labels = c("0-2000", "2000-5000", "5000-10000")))

@@

graph4_data = merged_bison_data |>
  filter(state_code == "OKLAHOMA")

@@

#Overall trends across the great plains... see downward trend overall
ggplot(graph2_data, aes(x = year, y = total_population)) + 
  xlab("Year") +
  ylab("Population by State") +
  labs(title = "Bison Population Over Time") +
  geom_point(aes(color = population_bin), size = 3) +
  geom_smooth()
#Each point on this plot represents a state for a given year, and the color represents the population bin that the state falls into.

#Number of Hunting licenses Over Time
ggplot(graph2_data, aes(x = year, y = `Paid Hunting License Holders`)) + 
  xlab("Year") +
  ylab("Number of Licenses") +
  labs(title = "Number of Licenses Over Time") +
  geom_point(aes(color = state_code), size = 3)
#Removed Texas from this plot because it was an extreme value and make the graph less interpretable by extending the y-axis so much. Texas saw a small increase over the last time period but was otherwise steady.

#Looking for relationship between number of hunting licenses and population
ggplot(merged_bison_data, aes(x = `Paid Hunting License Holders`, y = total_population)) +
  geom_point() + 
  geom_smooth() +
  xlab("Number of Lisense Holders") +
  ylab("Bison Population") +
  labs(title = "Number of Licenses vs Population")
#Not a clear relationship. With so many states there are too many competing factors at play to control for with our limited data.

#What about just Oklahoma since it is one of the only states with increasing numbers of licenses?
ggplot(graph4_data, aes(x = `Paid Hunting License Holders`, y = total_population)) +
  geom_point() + 
  xlab("Number of Lisense Holders") +
  ylab("Bison Population") +
  labs(title = "Number of Licenses vs Population for Oklahoma")
#We still have not inferred causality, but there seems to be a negative relationship now. The lack of variation in hunting licenses across other states makes regression extremely difficult, but Oklahoma could be a better case study.

### 2024-12-02 Draft Exploratory Analysis.Rmd
Add files via upload

  After realizing that there is some variety in the details of population growth or decline across states, we sought to confirm our suspicion that the overall population trend is negative. What we see in the following graph is a confirmation of that suspicion, where population sees a strong downward trend across the Great Plains:
![Great Plains Population Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Population Over Time.png)

  As a side note, our decision to restrict our focus to the Great Plains was based off of our spatial data which indicates to us that the majority of Bison herds are concentrated in these states, as shows in the following image:
![Bison Herds in the U.S.](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Range Map.png)

  Once we confirmed that population is indeed declining, we started exploring fluctuations in hunting license numbers as a possible correlate. The following graph shows the number of hunting licenses across the same years and the same states as our previous visual:
![Great Plains Hunting License Numbers Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Licenses Over Time Fixed.png)

From this graph, we do not see any obvious trends in hunting licenses over time across the Great Plains. The following graph confirms our suspician that there is not an obvious correlation between hunting licenses and Bison population overall (population fluctuates in both directions as the number of hunting licenses increases):
![Hunting License Numbers vs Population Over Time](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/License vs Population.png)

@@

From this visualization, we see that most states do not have significant fluctuations in their number of hunting licenses. However, Texas saw an upward trend in licenses from 2012 to 2017. The following plot explores the relationship between the increase in hunting licenses and population changes in Texas:
![License vs Population in Texas](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/License vs Population in Texas.png)

@@

>>>>

  After realizing that there is some variety in the details of population growth or decline across states, we sought to confirm our suspicion that the overall population trend is negative. What we see in the following graph is a confirmation of that suspicion, where population sees a strong downward trend across the Great Plains:
![Great Plains Population Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/bison pop over time by state and binned.png)
In this plot, each point on this plot represents a state for a given year, and the color represents the population bin that the state falls into.

  As a side note, our decision to restrict our focus to the Great Plains was based off of our spatial data which indicates to us that the majority of Bison herds are concentrated in these states, as shows in the following image:
![Bison Herds in the U.S.](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/States of interest and ranges map.png)

  Once we confirmed that population is indeed declining, we started exploring fluctuations in hunting license numbers as a possible correlate. The following graph shows the number of hunting licenses across the same years and the same states as our previous visual:
![Great Plains Hunting License Numbers Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/licenses over time 2.png)
Note that we removed texas from this graph. It was an extreme value which extended our y-axis greatly. We noted that Texas had some variation from 2010 to 2017 which is useful for us, however it was excluded from the graph in order to better demonstrate variation in other states and make the plots more interpretable.

  From this graph, we do not see any obvious trends in hunting licenses over time across most states in the Great Plains. The following graph confirms our suspicion that there is not an obvious correlation between hunting licenses and Bison population overall (population fluctuates in both directions as the number of hunting licenses increases):
![Hunting License Numbers vs Population Over Time](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/license vs population 2.png)

@@

From this visualization, we see that most states do not have significant variation in their number of hunting licenses. However, Texas saw an upward trend in licenses from 2012 to 2017. The following plot explores the relationship between the increase in hunting licenses and population changes in Texas:
![License vs Population in Texas](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/License vs Population in Texas.png)

@@

  The other state that saw some variation in the number of hunting licenses issued is Oklahoma, as shown below:
![License vs Population in Texas](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Oklahoma Hunting vs Pop.png)
Although we still lack sufficient evidence to infer causality between the two variables, Oklahoma at least demonstrates a negative relationship between hunting licenses and Bison population.

  We initially expected a negative relationship between hunting licenses and bison populations as a consequence of overhunting a common pool resource across all states. If states were to issue more licenses, we thought that this would ultimately mean that populations decrease as a consequence of more bison being killed by hunters. We believe that this relationship is not so obvious for two main reasons. First, the allocation of hunting licenses is heavily regulated at the state level in order to prevent a tragedy of the commons. Second, a state's capacity to maintain large Bison populations is dependent on the amount of land that the animals have to graze. Thus, an increase in the number of hunting licenses issued could in fact be an indicator of an increased amount of private hunting land. As the demand for Bison tags increases, it raises the opportunity cost of ranchers with land adjacent to bison ranges and incentives them to allow Bison to expand their territory. Thus, although more Bison are being hunted, there is more land available for additional growth.
